---
title: "Peer-graded Assignment: Prediction Assignment Writeup"
author: "Olusanya Oluwole"
date: "1/22/2021"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
The practical machine learning assignment intend to demonstrated how to 
integrate knowledge gain during the enter course course. 

*"Using devices such as 
Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large 
amount of data about personal activity relatively inexpensively. These type of 
devices are part of the quantified self movement – a group of enthusiasts who 
take measurements about themselves regularly to improve their health, to find 
patterns in their behavior, or because they are tech geeks. One thing that 
people regularly do is quantify how much of a particular activity they do, 
but they rarely quantify how well they do it."* The goal of this project is to 
use data from accelerometers on the belt, forearm, arm, and dumbell of 6 
participants to predict the manner in which they did the exercise (classe 
variable).

# Data Source Data Cleaning and Exploration
Our data is source from http://groupware.les.inf.puc-rio.br/har 

- Training: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
- Testing: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

```{r, warning=FALSE, message=FALSE}
#install.packages("parallel")
#install.packages("doParallel")
library(tidyverse)
library(caret)
library(rattle)
library(parallel)
library(doParallel)
trainUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
training <- read_csv(trainUrl)
testing <- read_csv(testUrl)
```

First, let remove near-zero variance predictor that will not contribute to 
our prediction from our training and testing data  using `nearZeroVar` fuction.
```{r}
# training dataset
nzvTraining <- nearZeroVar(training)
training <- training[, -nzvTraining]

# testing dataset
nzvtesting <- nearZeroVar(testing)
testing <- testing[, -nzvtesting]
```

There are still butch of `NAs` that can be removed as follow.
```{r}
training <- training[, colSums(is.na(training)) == 0]
testing <- testing[, colSums(is.na(testing)) == 0]
```

Variables `X1`, `user_name`, `raw_timestamp_part_1`, `raw_timestamp_part_2` and
`cvtd_timestamp` are not going to be useful for our prediction. These will be
remove as follow.
```{r}
training <- select(training, -c(1:5))
testing <- select(testing, -c(1:5))
dim(training); dim(testing)
```

## Exploratory Analysis with Correlation
Let take a look at how some of the variables are related to `classe` variable
with `featurePlot`.
```{r, message=FALSE, warning=FALSE}
featurePlot(x=training[, c("accel_belt_y", "accel_dumbbell_z", "accel_forearm_y",
                           "accel_arm_y", "accel_belt_z", "roll_belt", "gyros_arm_y")],
            y = training$classe,
            plot="pairs")
```
The panel show some variables are correlated while others are not.

The dataset `testing` will be set aside for answering the **Quiz Portion** of 
this project. The `training` dataset will be further divided to `trainingData` 
(70%) and `testingData` (30%) for the portion of the first portion of this 
projected.
```{r, message=FALSE, warning=FALSE}
inTrain <- createDataPartition(training$classe, p = 0.7, list = FALSE)
trainingData <- training[inTrain,]
testingData <- training[-inTrain,]
```

# Prediction Models
```{r}
intervalStart <- Sys.time()
cluster <- makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)
fitControl <- trainControl(method = "cv",
number = 5,
allowParallel = TRUE)
```

## Decision Tree
```{r, message=FALSE, warning=FALSE}
set.seed(12345)
treeMod <- train(classe ~ .,method="rpart",data=trainingData)
fancyRpartPlot(treeMod$finalModel)
treePredict <- predict(treeMod, newdata=testingData)
treeConfusionMatrix <- confusionMatrix(treePredict, factor(testingData$classe))
decisionTreeAccuracy <- treeConfusionMatrix$overall["Accuracy"]
```

## Generalized Boosted Model
```{r, message=FALSE, warning=FALSE, results='hide'}
set.seed(12345)
system.time(gbmMod <- train(classe ~ .,method="gbm",data=trainingData, trControl = fitControl))
gbmPredict <- predict(gbmMod, newdata=testingData)
gbmConfusionMatrix <- confusionMatrix(gbmPredict, factor(testingData$classe))
boostingAccuracy <- gbmConfusionMatrix$overall["Accuracy"]
```

## Random Forest
```{r, message=FALSE, warning=FALSE, results='hide'}
set.seed(12345)
system.time(rfMod <- train(classe ~ .,method="rf",data=trainingData, trControl = fitControl))
rfPredict <- predict(rfMod,newdata=testingData)
rfConfusionMatrix <- confusionMatrix(rfPredict, factor(testingData$classe))
randomForestAccuracy <- rfConfusionMatrix$overall["Accuracy"]
```

## Linear Discriminant Analysis (LDA)
```{r, message=FALSE, warning=FALSE, results='hide'}
set.seed(12345)
system.time(ldaMod <- train(classe ~ .,method="lda",data=trainingData, trControl = fitControl))
ldaPredict <- predict(ldaMod, newdata=testingData)
ldaConfusionMatrix <- confusionMatrix(ldaPredict, factor(testingData$classe))
ldaAccuracy <- ldaConfusionMatrix$overall["Accuracy"]
```

```{r, message=FALSE, warning=FALSE}
accuracyTable <- tibble(decisionTreeAccuracy, boostingAccuracy,
                        randomForestAccuracy, ldaAccuracy)
accuracyTable 
```
Of the four prediction method use on the training dataset, random forest is the
most accurate with score of approximately 99%.

# Class Prediction for `testing` Dataset
Random forest gives most accurate prediction on the test portion of the 
training dataset. Therefore, it will be use to predict class of testing
dataset that will be use to answer the course project quiz portion.
```{r, message=FALSE, warning=FALSE}
rfPredict2 <- predict(rfMod,newdata=testing)
rfPredict2
```

